CrypTool Script \\ Mathematics and Cryptography

### 2002-08-26 script/en/chapters/numbertheory.tex
be comments

\begin{center}
\fbox{\parbox{15cm}{{\em Carl Friedrich Gauss\index{Gauss} (30.4.1777 - 23.2.1855):}\\
Mathematics is the queen of sciences and number theory is the queen of mathematics.}}
\end{center}

%\begin{quote}
%{\em Carl Friedrich Gauss (30.4.1777 - 23.2.1855):}\\
%Mathematics is the queen of sciences and number theory is the queen of mathematics.
%\end{quote}

@@

Number theory arose from interest in positive whole numbers $1, 2, 3, 4, \cdots$, also referred to as the set of natural numbers 
\index{Number!natural} {\em natural numbers} $\mathbb{N}$. These are the first mathematical constructs 
used by human civilisation. According to Kronecker, they are a creation of God. 
In Dedekind's opinion, they are a creation of the human intellect. Dependent upon one's ideology,
this is an unsolvable contradiction or one and the same thing.

@@

If you allow numbers other than primes as factors, there are several ways of factorising integers and the {\em 
uniqueness} is lost:
$$60 = 1*60 = 2*30 = 4*15 = 5*12 = 6*10 = 2*3*10 = 2*5*6 = 3*4*5 = \cdots$$
The 1st fundamental theorem only appears to be obvious. We can construct numerous other sets of 
numbers\footnote{These sets are formed especially from the set of natural numbers. An example of this can be found in this \hyperlink{uniqueness}{script} on page 
\pageref{remFundTheoOfArithm} at the end of chapter \ref{primesinmath}} for which numbers in the set {\em cannot} be expressed 
uniquely as a product of the prime numbers of the set.

@@

From algebra theorems it follows that essential parts of the conventional calculation rules are kept when we 
proceed to modular calculations over a basic set $\mathbb{Z}$.  For example, addition remains commutative. 
The same goes for multiplication modulo $m$. The result of a division\footnote{\label{ftn-zth-divmodn}When dividing modulo $m$  we cannot use every number because some numbers have the same 
property as zero. See footnote \ref{ftn-zth-closed}.} is not a fraction but rather a whole number between $0$ and $m-1$.

@@

\begin{itemize}
\item the addition table\footnote{Comment on subtraction modulo 5: \\
$2 - 4 \equiv -2 \equiv 3{\rm ~mod~}5.$\\
It is therefore not true modulo $5$ that $-2 = 2$ (see also \hyperlink{Appendix_C}{Appendix C}). }
${\rm ~(mod~ } 5)$ and
\item the multiplication tables\footnote{Comment on division modulo $6$: 
Due to the special role of zero as the identity for addition, division by zero is not permitted:\\
for all $a$ it is $a*0=0,$ because $a*0=a*(0+0)=a*0+a*0.$ Obviously $0$ has no inverse regarding the
multiplication, because if there would be one, it must be $ 0= 0*0^{-1} =1.$ Also see footnote \ref{ftn-zth-divmodn}.}
for mod $5$ and mod $6$.
\end{itemize}
\subsubsection*{Example of an addition table}
The result when we add $3$ and $4 {\rm ~(mod~ } 5)$ is calculated as follows: Calculate $3 + 4 = 7$ and keep 
subtracting $5$ from the result until the result is less than the modulo:  $7 - 5 = 2$. 
Therefore:  $3 + 4 \equiv 2 {\rm ~(mod~ } 5)$.
%
\begin{table}[h]
\begin{center}
\begin{tabular}{r|ccccc}
Addition table modulo 5: \quad + &  0 & 1 & 2 & 3 & 4  \\
\hline
0 &  0 & 1 & 2 & 3 & 4 \\  
1 & 1 &  2 & 3 & 4 & 0 \\
2 & 2 & 3 & 4 & 0 & 1 \\
3 & 3 & 4 & 0 & 1 & 2 \\
4 & 4 & 0 & 1 & 2 & 3 \\[-5ex] % remove some extra space 
\end{tabular} 
\end{center} 
\end{table}
% ..............................................................................................
\subsubsection*{Example of an multiplication table:}
The result of the multiplication $4 * 4 {\rm ~(mod~ } 5)$ is calculated as follows: $4 * 4 = 
16$ and subtract $5$ until the result is less than the modulus.
$$16 - 5 = 11;~ 11 - 5 = 6;~6- 5 = 1.$$
The table directly shows that $4 * 4 \equiv 1 {\rm ~(mod~} 5)$ because $16 : 5 = 3$ remainder $1$. 
Multiplication is defined on the set $\mathbb{Z}$  excluding $0$.
%
\begin{table}[h]
\begin{center}
\begin{tabular}{r|cccc}
Multiplication table modulo 5: \quad * & 1& 2 & 3 & 4  \\
\hline 
1 & 1 &    2    &    3    & 4 \\
2 & 2 & {\bf 4} & {\bf 1} & 3 \\ 
3 & 3 & {\bf 1} & {\bf 4} & 2  \\
4 & 4 &    3    &    2    & 1 
\end{tabular}
\end{center} 
\end{table}

@@

\subsubsection{Fast calculation of high powers} \hypertarget{hohpot}{} \label{hohpot}\index{Power}
%\subsubsection*{3.6.3.1 ~~~Fast calculation of high powers} \hypertarget{hohpot}{} \index{Power}
%\addcontentsline{toc}{subsubsection}{~~~~~~~~~~3.6.3.1 ~Fast calculation of high powers}
RSA encryption and decryption\footnote{See \hyperlink{Kapitel_3_10}{chapter 3.10 Proof of the RSA procedure with Euler-Fermat} \index{RSA}
and \hyperlink{Kapitel_3_12}{chapter 3.12 The RSA procedure with actual numbers}.} entails calculating high powers modulo $m$. 
For example, the calculation ($100^5) {\rm ~(mod~} 3)$ exceeds the 32-bit long integer\index{Long integer} number range provided we calculate $a^n$ 
by actually multiplying a with itself $n$ times in line with the definition. In the case of extremely large numbers, even 
a fast computer chip would take longer than the age of the universe to calculate a single exponential. Luckily, there is 
an extremely effective shortcut for calculating exponentials (but not for calculating logarithms).

@@

\subsection{Multiplicative order and primitive roots}
\hypertarget{Kap_3_9}{} \ref{MultOrdPrimitveRoot}

@@

\subsubsection*{Table 6: Values of $a^i{\rm ~mod~}46, 1\leq a,i<23$:}\label{SrcArith3c}
{ \textmd \small
\begin{center}  
\begin{tabular}{|p{16 pt}||@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|c|}
\hline
$a \setminus i$   & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16 & 17 & 18 & 19 & 20 & 21 & 22 & 23 & ord \\
\hline
\hline
1    & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1 & 1    \\
\hline
2 & 2  & 4  & 8 & 16 & 32 & 18 & 36 & 26  & 6 & 12 & 24  & 2  & 4  & 8 & 16 & 32 & 18 & 36 & 26  & 6 & 12 & 24  & 2 & --    \\
\hline
3 & 3  & 9 & 27 & 35 & 13 & 39 & 25 & 29 & 41 & 31  & 1  & 3  & 9 & 27 & 35 & 13 & 39 & 25 & 29 & 41 & 31  & 1  & 3 & 11   \\
\hline
4  & 4 & 16 & 18 & 26 & 12  & 2  & 8 & 32 & 36  & 6 & 24  & 4 & 16 & 18 & 26 & 12  & 2  & 8 & 32 & 36  & 6 & 24  & 4 & --  \\
\hline
5 & 5 & 25 & 33 & 27 & 43 & 31 & 17 & 39 & 11  & 9 & 45 & 41 & 21 & 13 & 19  & 3 & 15 & 29  & 7 & 35 & 37  & 1  & 5 & 22  \\
\hline
6 & 6 & 36 & 32  & 8  & 2 & 12 & 26 & 18 & 16  & 4 & 24  & 6 & 36 & 32  & 8  & 2 & 12 & 26 & 18 & 16  & 4 & 24  & 6 & -- \\
\hline
7 & 7  & 3 & 21  & 9 & 17 & 27  & 5 & 35 & 15 & 13 & 45 & 39 & 43 & 25 & 37 & 29 & 19 & 41 & 11 & 31 & 33  & 1  & 7 & 22 \\
\hline
8 & 8 & 18  & 6  & 2 & 16 & 36 & 12  & 4 & 32 & 26 & 24  & 8 & 18  & 6  & 2 & 16 & 36 & 12  & 4 & 32 & 26 & 24  & 8 & --  \\
\hline
9 & 9 & 35 & 39 & 29 & 31  & 3 & 27 & 13 & 25 & 41  & 1  & 9 & 35 & 39 & 29 & 31  & 3 & 27 & 13 & 25 & 41  & 1  & 9 & 11  \\
\hline
10 & 10  & 8 & 34 & 18 & 42  & 6 & 14  & 2 & 20 & 16 & 22 & 36 & 38 & 12 & 28  & 4 & 40 & 32 & 44 & 26 & 30 & 24 & 10 & --  \\
\hline 
11 & 11 & 29 & 43 & 13  & 5  & 9  & 7 & 31 & 19 & 25 & 45 & 35 & 17  & 3 & 33 & 41 & 37 & 39 & 15 & 27 & 21  & 1 & 11 & 22 \\
\hline
12 & 12  & 6 & 26 & 36 & 18 & 32 & 16  & 8  & 4  & 2 & 24 & 12  & 6 & 26 & 36 & 18 & 32 & 16  & 8  & 4  & 2 & 24 & 12 & -- \\
\hline
13 & 13 & 31 & 35 & 41 & 27 & 29  & 9 & 25  & 3 & 39  & 1 & 13 & 31 & 35 & 41 & 27 & 29  & 9 & 25  & 3 & 39  & 1 & 13 & 11  \\
\hline
14 & 14 & 12 & 30  & 6 & 38 & 26 & 42 & 36 & 44 & 18 & 22 & 32 & 34 & 16 & 40  & 8 & 20  & 4 & 10  & 2 & 28 & 24 & 14 & -- \\
\hline
15 & 15 & 41 & 17 & 25  & 7 & 13 & 11 & 27 & 37  & 3 & 45 & 31  & 5 & 29 & 21 & 39 & 33 & 35 & 19  & 9 & 43  & 1 & 15 & 22 \\
\hline
16 & 16 & 26  & 2 & 32  & 6  & 4 & 18 & 12  & 8 & 36 & 24 & 16 & 26  & 2 & 32  & 6  & 4 & 18 & 12  & 8 & 36 & 24 & 16 & -- \\
\hline
17 & 17 & 13 & 37 & 31 & 21 & 35 & 43 & 41  & 7 & 27 & 45 & 29 & 33  & 9 & 15 & 25 & 11  & 3  & 5 & 39 & 19  & 1 & 17 & 22 \\
\hline
18 & 18  & 2 & 36  & 4 & 26  & 8  & 6 & 16 & 12 & 32 & 24 & 18  & 2 & 36  & 4 & 26  & 8  & 6 & 16 & 12 & 32 & 24 & 18 & -- \\
\hline
19  & 19 & 39  & 5  & 3 & 11 & 25 & 15  & 9 & 33 & 29 & 45 & 27  & 7 & 41 & 43 & 35 & 21 & 31 & 37 & 13 & 17  & 1 & 19 & 22 \\
\hline
20  & 20 & 32 & 42 & 12 & 10 & 16 & 44  & 6 & 28  & 8 & 22 & 26 & 14  & 4 & 34 & 36 & 30  & 2 & 40 & 18 & 38 & 24 & 20 & --  \\
\hline
21 & 21 & 27 & 15 & 39 & 37 & 41 & 33  & 3 & 17 & 35 & 45 & 25 & 19 & 31  & 7  & 9  & 5 & 13 & 43 & 29 & 11  & 1 & 21 & 22  \\
\hline
22 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & -- \\
\hline
23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & -- \\
\hline 
\end{tabular}
\end{center}
}

@@

\hypertarget{RSABeweis} Using the Euler-Fermat theorem, we can ``prove'' the
RSA\footnote{The RSA procedure is the most common asymmetric 
cryptography procedure. Developed in 1978 by Ronald Rivest, Adi Shamir and Leonard Adleman, it can be used both for
signatures and for encryption. Cryptographers always associate this procedure with the abbreviation ``RSA''. However, 
please note that the abbreviation may be used in other fields to mean
something completely different.}$^,$\footnote{%
Within literature and movies not only cassical methods are used (like the
secret message solved by Sherlock Holmes in ``The Adventure of the Dancing
Men'' by Arthur Conan  Doyle), but also modern schemes. Examples are
\begin{itemize}
\item the movie ``House of Cards'', 1992, where autistic children
  communicate via primes,
\item the story ``The Dialogue of the Sisters'' by Dr.~C.~Elsner, 1999
  (included in the CrypTool package as PDF file), where the sisters
  communicate confidentially using a variant of RSA.
\end{itemize}
} 
procedure in the group $\mathbb{Z}_n^*$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Considerations regarding the security of the RSA algorithm}\label{SecurityRSA}

Major parts of this chapter follow the article ``Vorz\"uge und Grenzen des
RSA-Verfahrens'' written by F.~Bourseau / D.~Fox / C.~Thiel, and published
in \emph{Datenschutz und Datensicherheit} (DuD) 26/2002, pp. 84$-$89.

There have always been discussions about the suitability of the RSA
algorithm for digital signatures and encryption, e.~g.\ after publications of
breakthroughs in factorization. Nevertheless the RSA algorithm has become a
de-facto standard since it was published more than 20 years ago.

@@

\subsubsection{Complexity}

@@

This formula shows, that the factorization problem belongs to the class of
problems with sub-exponential time complexity (i.~e.\ time complexity grows
asymptotically not as fast as exponential functions like $e^l$ or $2^l$,
but stricly slower, e.~g.\ like $e^{\sqrt{l}}$).  This classification is
all that is currently known; it does not preclude the possiblity that the
factorization problem can be solved in polynomial time (s.~\ref{RSABernstein}).

$O(l)$ is the average number of processor steps depending on the bit length
$l$ of the number $n$ to be factored.  For the best currently known
factorization algorithm the constant $c = (64/9)^/173) = 1,923$.

@@

Within the article ``Vorz\"uge und Grenzen des RSA-Verfahrens'' Dirk Fox\footnote{%
The company Secorvo GmbH made a statement on the recommendation for key
length selection published by the GISA\index{GISA} (German Information
Security Agency). Chapter 2.3.1 of their statement contains a very
competent and understandable discussion of RSA security (this document
exists to my
knowledge only in German):\\
\href{http://www.secorvo.de/publikat/stellungnahme-algorithmenempfehlung-020307.pdf}
{\texttt{http://www.secorvo.de/publikat/stellungnahme-algorithmenempfehlung-020307.pdf}}
}
prognosticates an almost linear progression with the factorization, if all
factors are included: Each year the module length feasible for
factorization increases by 20 bit on average.  Please refer: DuD, 26/2002,
pp. 84-89, F.~Bourseau / D.~Fox / C.~Thiel

@@

\begin{minipage}{\textwidth}
Links:

\vskip -10 pt
\begin{itemize}
  \item[] \href{http://cr.yp.to/djb.html}
               {\texttt{http://cr.yp.to/djb.html}} \\
          \href{http://www.counterpane.com/crypto-gram-0203.html\#6}
               {\texttt{http://www.counterpane.com/crypto-gram-0203.html\#6}} \\
          \href{http://www.math.uic.edu}
               {\texttt{http://www.math.uic.edu}}
\end{itemize}
\end{minipage}

@@

The 5 smaller prime factors, already found before have been:
$$
\begin{array}{c}
3, 1907, 425796183929, \\
1624700279478894385598779655842584377, \\
3802306738549441324432139091271828121 \quad {\rm and} \\
128064886830166671444802576129115872060027.
\end{array}
$$
The first 3 factors can be easily computed (e.g. using CrypTool via menu {\bf Indiv. Procedures / RSA Demonstration / Factorise}).
The remaining three factors were fond by P. Zimmerman
(\href{http://www.loria.fr/~zimmerma/ecmnet}{\tt http://www.loria.fr/~zimmerma/ecmnet}), 
T. Grandlund (\href{http://www.swox.se/gmp/}{\tt http://www.swox.se/gmp/})
and R. Harley during the years 1999 and 2000 using the elliptic curve factorisation Method.

>>>>

\begin{center}
\fbox{\parbox{15cm}{{\em Carl Friedrich Gauss\index{Gauss} (30.4.1777$-$23.2.1855):}\\
Mathematics is the queen of sciences and number theory is the queen of mathematics.}}
\end{center}

%\begin{quote}
%{\em Carl Friedrich Gauss (30.4.1777$-$23.2.1855):}\\
%Mathematics is the queen of sciences and number theory is the queen of mathematics.
%\end{quote}

@@

Number theory arose from interest in positive whole numbers $1, 2, 3, 4, \cdots$, also referred to as the set of natural numbers 
\index{Number!natural} {\em natural numbers} $\mathbb{N}$. These are the first mathematical constructs 
used by human civilisation. According to Kronecker\footnote{Leopold
  Kronecker, German mathematician, 1823$-$1891.}\index{Kronecker Leopold}, they are a creation of God. 
In Dedekind's\footnote{Julius Wilhelm Richard Dedekind,
  German mathematician, 06.10.1831$-$12.02.1916.}\index{Dedekind Julius} opinion, they are a creation of the human intellect. Dependent upon one's ideology,
this is an unsolvable contradiction or one and the same thing.

@@

If you allow numbers other than primes as factors, there are several ways of factorising integers and the {\em 
uniqueness} is lost:
$$60 = 1*60 = 2*30 = 4*15 = 5*12 = 6*10 = 2*3*10 = 2*5*6 = 3*4*5 = \cdots$$
The 1st fundamental theorem only appears to be obvious. We can construct numerous other sets of 
numbers\footnote{These sets are formed especially from the set of natural numbers. An example of this can be found in this \hyperlink{uniqueness}{script} on 
page~\pageref{thm-pz-euklid} % should be \pageref{remFundTheoOfArithm}, but
% hyperref seems to be buggy
at the end of chapter \ref{primesinmath}} for which numbers in the set {\em cannot} be expressed 
uniquely as a product of the prime numbers of the set.

@@

From algebra theorems it follows that essential parts of the conventional calculation rules are kept when we 
proceed to modular calculations over a basic set $\mathbb{Z}$.  For example, addition remains commutative. 
The same goes for multiplication modulo $m$. The result of a division\footnote{\label{ftn-zth-divmodn}When dividing modulo $m$\index{Division modulo $n$}  we cannot use every number because some numbers have the same 
property as zero. See footnote \ref{ftn-zth-closed}.} is not a fraction but rather a whole number between $0$ and $m-1$.

@@

\begin{itemize}
\item the addition table\footnote{Comment on subtraction modulo 5: \\
$2 - 4 \equiv -2 \equiv 3{\rm ~mod~}5.$\\
It is therefore not true modulo $5$ that $-2 = 2$ (see also \hyperlink{Appendix_C}{Appendix C}). }
${\rm ~(mod~ } 5)$ and
\item the multiplication tables\footnote{Comment on division modulo $6$:\index{Division modulo $n$}\\
Due to the special role of zero as the identity for addition, division by zero is not permitted:\\
for all $a$ it is $a*0=0,$ because $a*0=a*(0+0)=a*0+a*0.$ Obviously $0$ has no inverse regarding the
multiplication, because if there would be one, it must be $ 0= 0*0^{-1} =1.$ Also see footnote \ref{ftn-zth-divmodn}.}
for mod $5$ and mod $6$.
\end{itemize}
\subsubsection*{Example of an addition table}
The result when we add $3$ and $4 {\rm ~(mod~ } 5)$ is calculated as follows: Calculate $3 + 4 = 7$ and keep 
subtracting $5$ from the result until the result is less than the modulo:  $7 - 5 = 2$. 
Therefore:  $3 + 4 \equiv 2 {\rm ~(mod~ } 5)$.
%
\begin{table}[h]
\begin{center}
\begin{tabular}{r|ccccc}
Addition table modulo 5: \quad + &  0 & 1 & 2 & 3 & 4  \\
\hline
0 &  0 & 1 & 2 & 3 & 4 \\  
1 & 1 &  2 & 3 & 4 & 0 \\
2 & 2 & 3 & 4 & 0 & 1 \\
3 & 3 & 4 & 0 & 1 & 2 \\
4 & 4 & 0 & 1 & 2 & 3 \\[-5ex] % remove some extra space 
\end{tabular} 
\end{center} 
\end{table}
% ..............................................................................................
\subsubsection*{Example of an multiplication table:}
The result of the multiplication $4 * 4 {\rm ~(mod~ } 5)$ is calculated as follows: $4 * 4 = 
16$ and subtract $5$ until the result is less than the modulus.
$$16 - 5 = 11;~ 11 - 5 = 6;~6- 5 = 1.$$
The table directly shows that $4 * 4 \equiv 1 {\rm ~(mod~} 5)$ because $16 : 5 = 3$ remainder $1$. 
Multiplication is defined on the set $\mathbb{Z}$  excluding $0$.
%
\begin{table}[h]
\begin{center}
\begin{tabular}{r|cccc}
Multiplication table modulo 5: \quad * & 1& 2 & 3 & 4  \\
\hline 
1 & 1 &    2    &    3    & 4 \\
2 & 2 & {\bf 4} & {\bf 1} & 3 \\ 
3 & 3 & {\bf 1} & {\bf 4} & 2  \\
4 & 4 &    3    &    2    & 1 
\end{tabular}
\end{center} 
\end{table}

@@

\subsubsection{Fast calculation of high powers} \hypertarget{hohpot}{} \label{hohpot}\index{Power}
%\subsubsection*{3.6.3.1 ~~~Fast calculation of high powers} \hypertarget{hohpot}{} \index{Power}
%\addcontentsline{toc}{subsubsection}{~~~~~~~~~~3.6.3.1 ~Fast calculation of high powers}
RSA encryption and decryption\footnote{See \hyperlink{Kapitel_3_10}{chapter 3.10 Proof of the RSA procedure with Euler-Fermat} \index{RSA}
and \hyperlink{Kapitel_3_12}{chapter 3.13 The RSA procedure with actual numbers}.} entails calculating high powers modulo $m$. 
For example, the calculation ($100^5) {\rm ~(mod~} 3)$ exceeds the 32-bit long integer\index{Long integer} number range provided we calculate $a^n$ 
by actually multiplying a with itself $n$ times in line with the definition. In the case of extremely large numbers, even 
a fast computer chip would take longer than the age of the universe to calculate a single exponential. Luckily, there is 
an extremely effective shortcut for calculating exponentials (but not for calculating logarithms).

@@

\pagebreak
\subsection{Multiplicative order and primitive roots}
\hypertarget{Kap_3_9}{} \label{MultOrdPrimitveRoot}

@@

\subsubsection*{Table 6: Values of $a^i{\rm ~mod~}46, 1\leq a,i<23$:}\label{SrcArith3c}
{ \textmd \small
\begin{center}  
\begin{tabular}{|p{16 pt}||@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|c|}
\hline
$a \setminus i$   & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16 & 17 & 18 & 19 & 20 & 21 & 22 & 23 & ord \\
\hline
\hline
1    & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1 & 1    \\
\hline
2 & 2  & 4  & 8 & 16 & 32 & 18 & 36 & 26  & 6 & 12 & 24  & 2  & 4  & 8 & 16 & 32 & 18 & 36 & 26  & 6 & 12 & 24  & 2 & --    \\
\hline
3 & 3  & 9 & 27 & 35 & 13 & 39 & 25 & 29 & 41 & 31  & 1  & 3  & 9 & 27 & 35 & 13 & 39 & 25 & 29 & 41 & 31  & 1  & 3 & 11   \\
\hline
4  & 4 & 16 & 18 & 26 & 12  & 2  & 8 & 32 & 36  & 6 & 24  & 4 & 16 & 18 & 26 & 12  & 2  & 8 & 32 & 36  & 6 & 24  & 4 & --  \\
\hline
5 & 5 & 25 & 33 & 27 & 43 & 31 & 17 & 39 & 11  & 9 & 45 & 41 & 21 & 13 & 19  & 3 & 15 & 29  & 7 & 35 & 37  & 1  & 5 & 22  \\
\hline
6 & 6 & 36 & 32  & 8  & 2 & 12 & 26 & 18 & 16  & 4 & 24  & 6 & 36 & 32  & 8  & 2 & 12 & 26 & 18 & 16  & 4 & 24  & 6 & -- \\
\hline
7 & 7  & 3 & 21  & 9 & 17 & 27  & 5 & 35 & 15 & 13 & 45 & 39 & 43 & 25 & 37 & 29 & 19 & 41 & 11 & 31 & 33  & \textbf{1}  & 7 & 22 \\
\hline
8 & 8 & 18  & 6  & 2 & 16 & 36 & 12  & 4 & 32 & 26 & 24  & 8 & 18  & 6  & 2 & 16 & 36 & 12  & 4 & 32 & 26 & 24  & 8 & --  \\
\hline
9 & 9 & 35 & 39 & 29 & 31  & 3 & 27 & 13 & 25 & 41  & 1  & 9 & 35 & 39 & 29 & 31  & 3 & 27 & 13 & 25 & 41  & 1  & 9 & 11  \\
\hline
10 & 10  & 8 & 34 & 18 & 42  & 6 & 14  & 2 & 20 & 16 & 22 & 36 & 38 & 12 & 28  & 4 & 40 & 32 & 44 & 26 & 30 & 24 & 10 & --  \\
\hline 
11 & 11 & 29 & 43 & 13  & 5  & 9  & 7 & 31 & 19 & 25 & 45 & 35 & 17  & 3 & 33 & 41 & 37 & 39 & 15 & 27 & 21  & 1 & 11 & 22 \\
\hline
12 & 12  & 6 & 26 & 36 & 18 & 32 & 16  & 8  & 4  & 2 & 24 & 12  & 6 & 26 & 36 & 18 & 32 & 16  & 8  & 4  & 2 & 24 & 12 & -- \\
\hline
13 & 13 & 31 & 35 & 41 & 27 & 29  & 9 & 25  & 3 & 39  & 1 & 13 & 31 & 35 & 41 & 27 & 29  & 9 & 25  & 3 & 39  & 1 & 13 & 11  \\
\hline
14 & 14 & 12 & 30  & 6 & 38 & 26 & 42 & 36 & 44 & 18 & 22 & 32 & 34 & 16 & 40  & 8 & 20  & 4 & 10  & 2 & 28 & 24 & 14 & -- \\
\hline
15 & 15 & 41 & 17 & 25  & 7 & 13 & 11 & 27 & 37  & 3 & 45 & 31  & 5 & 29 & 21 & 39 & 33 & 35 & 19  & 9 & 43  & 1 & 15 & 22 \\
\hline
16 & 16 & 26  & 2 & 32  & 6  & 4 & 18 & 12  & 8 & 36 & 24 & 16 & 26  & 2 & 32  & 6  & 4 & 18 & 12  & 8 & 36 & 24 & 16 & -- \\
\hline
17 & 17 & 13 & 37 & 31 & 21 & 35 & 43 & 41  & 7 & 27 & 45 & 29 & 33  & 9 & 15 & 25 & 11  & 3  & 5 & 39 & 19  & 1 & 17 & 22 \\
\hline
18 & 18  & 2 & 36  & 4 & 26  & 8  & 6 & 16 & 12 & 32 & 24 & 18  & 2 & 36  & 4 & 26  & 8  & 6 & 16 & 12 & 32 & 24 & 18 & -- \\
\hline
19  & 19 & 39  & 5  & 3 & 11 & 25 & 15  & 9 & 33 & 29 & 45 & 27  & 7 & 41 & 43 & 35 & 21 & 31 & 37 & 13 & 17  & 1 & 19 & 22 \\
\hline
20  & 20 & 32 & 42 & 12 & 10 & 16 & 44  & 6 & 28  & 8 & 22 & 26 & 14  & 4 & 34 & 36 & 30  & 2 & 40 & 18 & 38 & 24 & 20 & --  \\
\hline
21 & 21 & 27 & 15 & 39 & 37 & 41 & 33  & 3 & 17 & 35 & 45 & 25 & 19 & 31  & 7  & 9  & 5 & 13 & 43 & 29 & 11  & 1 & 21 & 22  \\
\hline
22 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & -- \\
\hline
23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & -- \\
\hline 
\end{tabular}
\end{center}
}

@@

\hypertarget{RSABeweis} Using the Euler-Fermat theorem, we can ``prove'' the
RSA\footnote{The RSA procedure\index{RSA procedure} is the most common asymmetric 
cryptography procedure. Developed in 1978 by Ronald Rivest, Adi Shamir and Leonard Adleman, it can be used both for
signatures and for encryption. Cryptographers always associate this procedure with the abbreviation ``RSA''. However, 
please note that the abbreviation may be used in other fields to mean
something completely different.}$^,$\footnote{%
Within literature and movies not only cassical methods are used (like the
secret message solved by Sherlock Holmes in ``The Adventure of the Dancing
Men'' by Arthur Conan  Doyle), but also modern schemes. Examples are
\begin{itemize}
\item the movie ``House of Cards'', 1992, where autistic children
  communicate via primes,
\item the story ``The Dialogue of the Sisters'' by Dr.~C.~Elsner, 1999
  (included in the CrypTool package as PDF file), where the sisters
  communicate confidentially using a variant of RSA.
\end{itemize}
} 
procedure in the group $\mathbb{Z}_n^*$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection[Considerations regarding the security of the RSA algorithm]
{Considerations regarding the security of the RSA algorithm\footnote{Major
    parts of chapters \ref{complexity} and \ref{chptSecurityParam} follow
    the article ``Vorz\"uge und Grenzen des RSA-Verfahrens'' written by
    F.~Bourseau / D.~Fox / C.~Thiel, and published in \emph{Datenschutz und
      Datensicherheit} (DuD) 26/2002, pp.  84$-$89.}}\label{SecurityRSA}

There have always been discussions about the suitability of the RSA
algorithm for digital signatures and encryption, e.~g.\ after publications
of breakthroughs in factorization. Nevertheless the RSA algorithm has
become a de-facto standard since it was published more than 20 years ago
(compare~\ref{ECAlternative}).

@@

\subsubsection{Complexity}\label{complexity}\index{Complexity}

@@

This formula shows, that the factorization problem belongs to the class of
problems with sub-exponential time complexity (i.~e.\ time complexity grows
asymptotically not as fast as exponential functions like $e^l$ or $2^l$,
but strictly slower, e.~g.\ like $e^{\sqrt{l}}$).  This classification is
all that is currently known; it does not preclude the possiblity that the
factorization problem can be solved in polynomial time (see~\ref{RSABernstein}).

$O(l)$ is the average number of processor steps depending on the bit length
$l$ of the number $n$ to be factorized.  For the best currently known
factorization algorithm the constant $c = (64/9)^/173) = 1,923$.

@@

Within the article ``Vorz\"uge und Grenzen des RSA-Verfahrens'' Dirk Fox\footnote{%
The company Secorvo GmbH made a statement on the recommendation for key
length selection published by the GISA\index{GISA} (German Information
Security Agency). Chapter 2.3.1 of their statement contains a very
competent and understandable discussion of RSA security (this document
exists to my
knowledge only in German):\\
\href{http://www.secorvo.de/publikat/stellungnahme-algorithmenempfehlung-020307.pdf}
{\texttt{http://www.secorvo.de/publikat/stellungnahme-algorithmenempfehlung-020307.pdf}}
}
prognosticates an almost linear progression with the factorization, if all
factors are included: Each year the module length feasible for
factorization increases by 20 bit on average.

@@

\begin{minipage}{\textwidth}
Links:
\vspace{-10pt}
\begin{itemize}
  \item[] \href{http://cr.yp.to/djb.html}
               {\texttt{http://cr.yp.to/djb.html}} \\
          \href{http://www.counterpane.com/crypto-gram-0203.html\#6}
               {\texttt{http://www.counterpane.com/crypto-gram-0203.html\#6}} \\
          \href{http://www.math.uic.edu}
               {\texttt{http://www.math.uic.edu}}
\end{itemize}
\end{minipage}

@@

The 6 smaller prime factors, already found before have been:
$$
\begin{array}{c}
3, 1907, 425796183929, \\
1624700279478894385598779655842584377, \\
3802306738549441324432139091271828121 \quad {\rm and} \\
128064886830166671444802576129115872060027.
\end{array}
$$
The first 3 factors can be easily computed (e.g. using CrypTool via menu {\bf Indiv. Procedures / RSA Demonstration / Factorise}).
The next three prime factors were found by P.~Zimmerman
(\href{http://www.loria.fr/~zimmerma/ecmnet}{\tt http://www.loria.fr/~zimmerma/ecmnet}), 
T.~Grandlund (\href{http://www.swox.se/gmp/}{\tt http://www.swox.se/gmp/})
and R. Harley during the years 1999 and 2000 using the elliptic curve factorisation method.
